Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-12-13 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex

\author{
Hanne I. Oberman\\Utrecht University \And Johanna Munoz Avila\\Julius
Center for Health Sciences and Primary Care, University Medical Center
Utrecht, Utrecht University, Utrecht, The Netherlands \And Thomas P. A.
Debray\\Julius Center for Health Sciences and Primary Care, University
Medical Center Utrecht, Utrecht University, Utrecht, The
Netherlands \And Gerko Vink\\Utrecht University \And Valentijn M. T. de
Jong\\Julius Center for Health Sciences and Primary Care, University
Medical Center Utrecht, Utrecht University, Utrecht, The Netherlands
\AND Data Analytics and Methods Task Force, European Medicines Agency,
Amsterdam, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

@@

\Address{
    Hanne I. Oberman\\
    Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
          }

@@

When clustered datasets are affected by missing values, we can
distinguish between two types of missing data: sporadic missingness and
systematic missingness \citep{resc13}. Sporadic missingness arises when
variables are missing for some but not all of the units in a cluster
\citep{buur18, jola18}. For example, it is possible that test results
are missing for several students in one or more classes. {[}TODO:
Provide an example for one of the case studies below.{]} When all
observations are missing within one or more clusters, data are
systematically missing. {[}TODO: Refer to Figure 1 and put
interpretation in the figure caption.{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(miceadds)     # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel pooling
\end{CodeInput}
\end{CodeChunk}

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.986     0.194    25.704     4.917     0.000     9.201     0.927 
sex             0.818     0.357     2.290     4.130     0.082    62.169     0.989

                            Estimate 
Intercept~~Intercept|school    0.272 
Residual~~Residual             0.975 
ICC|school                     0.227

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.186     0.254    20.457     4.706     0.000    11.816     0.942 
sex             0.760     0.219     3.479     4.609     0.020    13.627     0.950

                            Estimate 
Intercept~~Intercept|school    0.345 
Residual~~Residual             1.578 
ICC|school                     0.182

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.000     0.356    14.027     4.283     0.000    28.723     0.976 
sex             1.056     0.465     2.270     4.105     0.084    76.708     0.991

                            Estimate 
Intercept~~Intercept|school    0.294 
Residual~~Residual             1.347 
ICC|school                     0.187

@@

The analysis model for this dataset is a prediction model with
\texttt{mort} as the outcome. In this tutorial we'll estimate the
adjusted prognostic effect of \texttt{ct} on unfortunate outcomes. The
estimand is the adjusted odds ratio for \texttt{ct}, after including
\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
into the analysis model:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat, rotate = TRUE)  # plot missingness pattern
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   3.00e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.56e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

As we can see\ldots{} {[}TODO: fill in.{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term    estimate   std.error  statistic      p.value
1             (Intercept) -2.33931814 0.337744863  -6.926288 4.565459e-12
2                    type -0.41950933 0.179122441  -2.342026 1.919717e-02
3                     age  0.03042682 0.001565356  19.437637 0.000000e+00
4 as.factor(motor_score)2 -0.66613691 0.069302767  -9.611981 0.000000e+00
5 as.factor(motor_score)3 -1.04915293 0.070119780 -14.962297 0.000000e+00
6 as.factor(motor_score)4 -1.50206863 0.071907169 -20.888997 0.000000e+00
7                   pupil  0.49317740 0.035150534  14.030438 0.000000e+00
8                      ct  0.42802910 0.029616105  14.452579 0.000000e+00
             b         df dfcom          fmi       lambda m          riv
1 2.240527e-04 10819.9749 11013 0.0025413258 0.0023569693 5 0.0023625377
2 2.066282e-05 10984.4464 11013 0.0009546917 0.0007728065 5 0.0007734042
3 1.927783e-08  8774.5121 11013 0.0096665946 0.0094408908 5 0.0095308707
4 1.096200e-04  3559.8620 11013 0.0279345878 0.0273886158 5 0.0281598759
5 3.652274e-05  8968.6823 11013 0.0091347491 0.0089138126 5 0.0089939832
6 9.004098e-05  4952.3514 11013 0.0212918407 0.0208966706 5 0.0213426612
7 1.183186e-04   293.7825 11013 0.1208778877 0.1149133407 5 0.1298328695
8 6.936809e-05   425.1630 11013 0.0991319347 0.0949041246 5 0.1048553276
          ubar
1 1.138027e-01
2 3.206005e-02
3 2.427206e-06
4 4.671329e-03
5 4.872956e-03
6 5.062592e-03
7 1.093578e-03
8 7.938720e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10509.778 10575.547 -5245.889 10491.778     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2939  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
                -2.3395                  -0.4185                   0.0304  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
                -0.6597                  -1.0504                  -1.5006  
                  pupil                       ct  
                 0.4893                   0.4302  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10501.337 10567.106 -5241.668 10483.337     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2896  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35407                 -0.42416                  0.03023  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66302                 -1.03875                 -1.50219  
                  pupil                       ct  
                0.50367                  0.43246  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10500.594 10566.363 -5241.297 10482.594     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2902  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.35434                 -0.41988                  0.03042  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65515                 -1.04987                 -1.49095  
                  pupil                       ct  
                0.49221                  0.43460  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10520.673 10586.442 -5251.336 10502.673     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2891  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32285                 -0.42261                  0.03062  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67111                 -1.05287                 -1.49941  
                  pupil                       ct  
                0.50321                  0.41360  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10532.391 10598.160 -5257.196 10514.391     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2866  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.32583                 -0.41243                  0.03045  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.68169                 -1.05384                 -1.51715  
                  pupil                       ct  
                0.47746                  0.42928  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

R=region+language+(1\textbar InterviewID) model with with interviewer as
random effects, because the observations are not independent. Interviews
are not allocated randomly. In theory we expect the
inclusion-restriction variable to be randomly assigned, that's why we're
adding region and language to compensate for non-random allocation.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("hiv",  package = "GJRM")
R> # We select  5 predictor variables over 9 regions
R> colnames(hiv)
\end{CodeInput}
\begin{CodeOutput}
 [1] "hivconsent"      "hiv"             "age"             "education"      
 [5] "wealth"          "region"          "marital"         "std"            
 [9] "age1sex_cat"     "highhiv"         "partner"         "condom"         
[13] "aidscare"        "knowsdiedofaids" "evertestedHIV"   "smoke"          
[17] "religion"        "ethnicity"       "language"        "interviewerID"  
[21] "sw"             
\end{CodeOutput}
\begin{CodeInput}
R> hivdata <- hiv[,c("hiv","hivconsent","age","marital","condom","highhiv","interviewerID","region","language")]
\end{CodeInput}
\end{CodeChunk}

Recode variables:

\begin{CodeChunk}
\begin{CodeInput}
R> # Study/group variable has to be recoded as integer
R> hivdata$region<-as.integer(hivdata$region)
R> 
R> # Categorical variables have to be recoded as factor
R> # to use the 2l.binary imputation method, it is required that the level names should not have any space between words
R> 
R> hivdata$hiv <- as.factor(hivdata$hiv) #to use the 2l.heckman method, it is required that
R> #the missing variable is stored as a factor in the dataset, otherwise the method will 
R> #apply the imputation correction for a  missing continuous variable instead for 
R> #a missing binary variable,which is in this case the binary response of the hiv test.
R> 
R> hivdata$marital <- as.factor(hivdata$marital)
R> levels(hivdata$marital)<-c("never_married","currently_married","formerly_married")
R> hivdata$condom <- as.factor(hivdata$condom)
R> levels(hivdata$condom)<-c("No_Condom_Last_Intercourse","Condom_Last_Intercourse")
R> hivdata$highhiv <- as.factor(hivdata$highhiv)
R> levels(hivdata$highhiv)<-c("Not_High_Risk_of_HIV","High_Risk_of_HIV")
R> hivdata$interviewerID <- as.factor(hivdata$interviewerID)
R> 
R> hivdata$interviewerID <-  as.factor(as.character(hivdata$interviewerID))
R> interv<-as.data.frame(table(hivdata$interviewerID,hivdata$region))
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> hivdata$hivconsent <- as.factor(hivdata$hivconsent)
R> ID_mixed <- lme4::glmer(hivconsent ~ region + language+(1 | interviewerID), data = hivdata, family = binomial)
R> reffect  <- ranef(ID_mixed)$interviewerID
R> reffect$interviewerID <- levels(hivdata$interviewerID)
R> colnames(reffect)     <- c("IDreffect","interviewerID")
R> hivdata  <-  merge(hivdata, reffect, by="interviewerID", all.x=TRUE)
R> 
R> hivdata$interviewerID<-NULL
R> hivdata$hivconsent<-NULL
R> hivdata$language<-NULL
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> #Set prediction matrix and methods
R> ini <- mice(hivdata, maxit = 0)
R> meth<-ini$method
R> meth["hiv"]<-"2l.binary"
R> pred <- ini$pred
R> pred[,"region"] <- 0
R> pred["region",] <- 0
R> pred["hiv","region"]<- -2
R> pred["hiv","IDreffect"]<- 0
R> 
R> # Heckman model
R> pred["hiv","IDreffect"] <- -3
R> meth<-ini$method
R> meth["hiv"]<-"2l.heckman"
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice( hivdata, # dataset with missing values
+                         m = 10,   # number of imputations
+                         maxit = 1,
+                         seed = 1234, #seed attached to the dataID
+                         meth = meth, #imputation method vector
+                         pred = pred, #imputation predictors matrix
+                         print = T,
+                         meta_method="reml",
+                         pmm=FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

>>>>

\author{
Hanne I. Oberman\\Utrecht University \And Johanna Munoz Avila\\Julius
Center for Health Sciences and Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands \AND Thomas P. A. Debray\\Julius Center
for Health Sciences and Primary Care \And Gerko Vink\\Utrecht University
\AND Valentijn M. T. de Jong\\Julius Center for Health Sciences and
Primary Care,\\
University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands
}
\title{Imputation of Incomplete Multilevel Data with \pkg{mice}}

@@

\Address{
    Hanne I. Oberman\\
    Utrecht University\\
    Padualaan 14\\
3584 CH Utrecht\\
  E-mail: \email{h.i.oberman@uu.nl}\\
  URL: \url{https://hanneoberman.github.io/}\\~\\
        Thomas P. A. Debray\\
    Julius Center for Health Sciences and Primary Care\\
    University Medical Center Utrecht, Utrecht University,\\
Utrecht, The Netherlands\\

@@

When clustered datasets are affected by missing values, we can
distinguish between two types of missing data: sporadic missingness and
systematic missingness \citep{resc13}. Sporadic missingness arises when
variables are missing for some but not all of the units in a cluster
\citep{buur18, jola18}. For example, it is possible that test results
are missing for several students in one or more classes. {[}TODO:
Provide an example for one of the case studies below.{]} When all
observations are missing within one or more clusters, data are
systematically missing. {[}TODO: Refer to Figure 1 and put
interpretation in the figure caption. VMTdJ: ``Mention in text that this
is sporadic missingness?''{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> set.seed(2022)
R> library(mice)         # for imputation
R> library(miceadds)     # for imputation
R> library(ggmice)       # for visualization
R> library(ggplot2)      # for visualization
R> library(dplyr)        # for data wrangling
R> library(lme4)         # for multilevel modeling
R> library(mitml)        # for multilevel pooling
R> library(miceheckman)  # for imputation
\end{CodeInput}
\end{CodeChunk}

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     5.012     0.295    16.994     4.362     0.000    22.587     0.969 
sex             0.695     0.251     2.768     4.287     0.047    28.390     0.975

                            Estimate 
Intercept~~Intercept|school    0.266 
Residual~~Residual             1.035 
ICC|school                     0.208

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902

                            Estimate 
Intercept~~Intercept|school    0.324 
Residual~~Residual             1.130 
ICC|school                     0.226

@@

             Estimate Std.Error   t.value        df   P(>|t|)       RIV       FMI 
(Intercept)     4.808     0.197    24.354     5.072     0.000     7.930     0.916 
sex             1.086     0.135     8.056     5.271     0.000     6.761     0.902

                            Estimate 
Intercept~~Intercept|school    0.324 
Residual~~Residual             1.130 
ICC|school                     0.226

@@

The analysis model for this dataset is a prediction model with
\texttt{mort} as the outcome. In this tutorial we'll estimate the
adjusted prognostic effect of \texttt{ct} on mortality outcomes. The
estimand is the adjusted odds ratio for \texttt{ct}, after including
\texttt{type}, \texttt{age} \texttt{motor\_score} and \texttt{pupil}
into the analysis model:

@@

\begin{CodeChunk}
\begin{CodeInput}
R> plot_pattern(dat, rotate = TRUE, npat = 10L)  # plot missingness pattern
\end{CodeInput}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- glmer(mod, family = "binomial", data = na.omit(dat)) # fit the model
R> tidy(fit, conf.int = TRUE, exponentiate = TRUE)             # print estimates
\end{CodeInput}
\begin{CodeOutput}
# A tibble: 11 x 9
   effect   group term        estim~1 std.er~2 stati~3   p.value conf.~4 conf.~5
   <chr>    <chr> <chr>         <dbl>    <dbl>   <dbl>     <dbl>   <dbl>   <dbl>
 1 fixed    <NA>  (Intercept)  0.0863  0.0182   -11.6   2.99e-31  0.0571   0.130
 2 fixed    <NA>  typeRCT      0.757   0.137     -1.54  1.22e- 1  0.531    1.08 
 3 fixed    <NA>  age          1.03    0.00265   12.9   7.40e-38  1.03     1.04 
 4 fixed    <NA>  motor_scor~  0.651   0.0732    -3.82  1.34e- 4  0.522    0.811
 5 fixed    <NA>  motor_scor~  0.489   0.0555    -6.30  2.97e-10  0.391    0.611
 6 fixed    <NA>  motor_scor~  0.274   0.0321   -11.0   2.28e-28  0.218    0.345
 7 fixed    <NA>  pupilNone    3.20    0.317     11.7   8.18e-32  2.63     3.88 
 8 fixed    <NA>  pupilOne     1.75    0.195      5.06  4.27e- 7  1.41     2.18 
 9 fixed    <NA>  ctIII        2.41    0.268      7.89  3.05e-15  1.94     2.99 
10 fixed    <NA>  ctIV/V       2.30    0.214      8.95  3.55e-19  1.92     2.76 
11 ran_pars name  sd__(Inter~  0.230  NA         NA    NA        NA       NA    
# ... with abbreviated variable names 1: estimate, 2: std.error, 3: statistic,
#   4: conf.low, 5: conf.high
\end{CodeOutput}
\end{CodeChunk}

As we can see\ldots{} {[}TODO: fill in: This means that a higher ct
(Marshall Computerized Tomography classification) is associated with a
lower odds of 6-month mortality, given by the odds ratio exp(0.42), CI
\ldots{} to \ldots, when controlling for \ldots{]}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> fit <- imp %>% 
+   with(glmer(mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 | name), family = "binomial")) 
R> tidy(pool(fit))
\end{CodeInput}
\begin{CodeOutput}
                     term   estimate   std.error  statistic      p.value
1             (Intercept) -2.3054756 0.340854445  -6.763813 1.425704e-11
2                    type -0.4200458 0.180837517  -2.322780 2.021049e-02
3                     age  0.0302518 0.001561625  19.372002 0.000000e+00
4 as.factor(motor_score)2 -0.6619736 0.069527977  -9.520967 0.000000e+00
5 as.factor(motor_score)3 -1.0520025 0.070748986 -14.869507 0.000000e+00
6 as.factor(motor_score)4 -1.5087772 0.072240455 -20.885489 0.000000e+00
7                   pupil  0.4731045 0.038187004  12.389150 0.000000e+00
8                      ct  0.4291135 0.032126389  13.357041 0.000000e+00
             b          df dfcom         fmi      lambda m         riv
1 7.937082e-04  9227.60708 11013 0.008412824 0.008197930 5 0.008265691
2 1.386594e-04 10229.67142 11013 0.005282534 0.005088076 5 0.005114097
3 1.138007e-08 10083.78177 11013 0.005796971 0.005599802 5 0.005631337
4 1.427540e-04  2450.43742 11013 0.036222753 0.035436457 5 0.036738334
5 1.147119e-04  3540.28145 11013 0.028049975 0.027501049 5 0.028278744
6 1.377409e-04  2902.11217 11013 0.032339200 0.031672563 5 0.032708526
7 3.160434e-04    58.71181 11013 0.284053948 0.260073900 5 0.351486317
8 1.985834e-04    74.37520 11013 0.250767587 0.230887509 5 0.300199921
          ubar
1 1.152293e-01
2 3.253582e-02
3 2.425016e-06
4 4.662835e-03
5 4.867765e-03
6 5.053394e-03
7 1.078995e-03
8 7.938047e-04
\end{CodeOutput}
\begin{CodeInput}
R> as.mitml.result(fit)
\end{CodeInput}
\begin{CodeOutput}
[[1]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10494.825 10560.593 -5238.412 10476.825     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2917  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33693                 -0.43143                  0.03013  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.67764                 -1.05120                 -1.49989  
                  pupil                       ct  
                0.49336                  0.44123  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[2]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10543.607 10609.376 -5262.803 10525.607     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2937  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.28159                 -0.42088                  0.03026  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65610                 -1.04928                 -1.49432  
                  pupil                       ct  
                0.48055                  0.41014  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[3]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10543.170 10608.939 -5262.585 10525.170     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2901  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.29948                 -0.40181                  0.03038  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.65959                 -1.06391                 -1.52075  
                  pupil                       ct  
                0.45837                  0.42186  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 2 lme4 warnings

[[4]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10527.222 10592.991 -5254.611 10509.222     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2962  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.27660                 -0.42912                  0.03016  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.66963                 -1.05952                 -1.51971  
                  pupil                       ct  
                0.45085                  0.44439  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

[[5]]
Generalized linear mixed model fit by maximum likelihood (Laplace
  Approximation) [glmerMod]
 Family: binomial  ( logit )
Formula: mort ~ type + age + as.factor(motor_score) + pupil + ct + (1 |  
    name)
      AIC       BIC    logLik  deviance  df.resid 
10519.178 10584.947 -5250.589 10501.178     11013 
Random effects:
 Groups Name        Std.Dev.
 name   (Intercept) 0.2897  
Number of obs: 11022, groups:  name, 15
Fixed Effects:
            (Intercept)                     type                      age  
               -2.33278                 -0.41699                  0.03032  
as.factor(motor_score)2  as.factor(motor_score)3  as.factor(motor_score)4  
               -0.64691                 -1.03611                 -1.50921  
                  pupil                       ct  
                0.48240                  0.42794  
optimizer (Nelder_Mead) convergence code: 0 (OK) ; 0 optimizer warnings; 1 lme4 warnings

@@

R=region+language+(1\textbar InterviewID) model with interviewer as
random effects, because the observations are not independent. Interviews
are not allocated randomly. In theory we expect the
inclusion-restriction variable to be randomly assigned, that's why we're
adding region and language to compensate for non-random allocation.
{[}TODO: Explain exclusion-restriction, refer to paper Johanna{]}.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> data("obesity",  package = "miceheckman")
\end{CodeInput}
\end{CodeChunk}

@@

@@

\begin{CodeChunk}
\begin{CodeInput}
R> #Set prediction matrix and methods
R> ini <- mice(obesity, maxit = 0)
R> meth<-ini$method
R> meth["weight"]<-"2l.heckman"
R> pred <- ini$pred
R> pred["weight","cluster"]<- -2
R> pred["weight","rt"] <- -3
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> imp <- mice(obesity, # dataset with missing values
+                         m = 10,   # number of imputations
+                         maxit = 1,
+                         seed = 1234, #seed attached to the dataID
+                         meth = meth, #imputation method vector
+                         pred = pred, #imputation predictors matrix
+                         print = T,
+                         meta_method="reml",
+                         pmm=FALSE)
\end{CodeInput}
\end{CodeChunk}

@@

\hypertarget{funding}{%
\section{Funding}\label{funding}}

This project has received funding from the European Union's Horizon 2020
research and innovation programme under ReCoDID grant agreement No
825746.

The views expressed in this paper are the personal views of the authors
and may not be understood or quoted as being made on behalf of or
reflecting the position of the regulatory agency/agencies or
organizations with which the authors are employed/affiliated.
